Iterative Monomorphisation

### 2024-03-22 iterative_monomorphisation.tex
added reference and added mangle as a primitve

\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage[noend]
   {algorithm2e}
\usepackage{amsmath}
\usepackage{enumitem}

@@

The initial phase of each iteration consists of computing two maps, \(M\) and \(N\), from the input problem~$\Phi$.
%
\begin{enumerate}
\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\). For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).

\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable. For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in N(\varphi)(\sym{foldl}) \).
\end{enumerate}

@@

The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop explosive enumerations.
Second, type arguments are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This dispenses from having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are complete, in a separate, final step.

The data structures used in the pseudocode are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, we keep track of a set \(S\) of substitutions generated by the matchings. It is used to generate new formulae in the final phase.

All sets referenced in the pseudocode are assumed to be finite. Additionally, this algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. They will therefore not be expanded upon. Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) is the substitution resulting from the matching. The pseudocode matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)}, where \(\sigma\) is a substitution, returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)}, where \(\sigma_1\) and \(\sigma_2\) are substitutions, tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
\end{enumerate}

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase
mangles composite types, so that we can target a simply typed logic with no general support for $n$-ary type constructors.
% \cite[Sections 4 and 5]{mono-trans}.

@@

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the monomorphisation process. It computes new type arguments for a single formula. A set of substitutions obtained from matching type argument tuples is iterated over. 
The separation of type arguments into old and new maps is put to use to ensure that only combinations involving at least one new map are considered. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings. Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.

The total number of type argument tuples can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth. If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

Similar bounds are used for the non-monomorphic type arguments:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
%This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration. Conversely, new type arguments can be limited per symbol, per formula, per iteration.

@@

The various bounds presented here overlap to some extent. For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its uses. For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.

@@

The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments are generated respectively for each symbol at each iteration. These options are closely linked and must be set together if they are to be relevant.

During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \cdot \text{number of type arguments})) \)

>>>>

\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage[noend]{algorithm2e}
\usepackage{amsmath}
\usepackage{enumitem}

@@

The initial phase of each iteration consists of computing two maps, \(M\) and \(N\), from the input problem~$\Phi$.
%
\begin{enumerate}
\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all monomorphic type arguments tuples to which \(f\) is applied in \(\Phi\). For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).

\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all type arguments tuples to which \(f\) is applied in \(\varphi\) and which contains a type variable. For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in N(\varphi)(\sym{foldl}) \).
\end{enumerate}

@@

The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop explosive enumerations.
Second, type arguments tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This dispenses from having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are complete, in a separate, final step.

The data structures used in the pseudocode are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments tuples having already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, we keep track of a set \(S\) of substitutions generated by the matchings. It is used to generate new formulae in the final phase.

All sets referenced in the pseudocode are assumed to be finite. Additionally, this algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. They will therefore not be expanded upon. Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) is the substitution resulting from the matching. The pseudocode matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)}, where \(\sigma\) is a substitution, returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)}, where \(\sigma_1\) and \(\sigma_2\) are substitutions, tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae returns the same set of formulae where all types have been mangled.
\end{enumerate}

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no general support for $n$-ary type constructors \cite[Sections 4 and 5]{monotonicity}.
% \cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the monomorphisation process. It computes new type arguments tuples for a single formula. A set of substitutions obtained from matching type argument tuples is iterated over. 
The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings. Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.

The total number of type argument tuples can increase cubicly in the number of type arguments tuples at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth. If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments tuples, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type arguments tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

Similar bounds are used for the non-monomorphic type arguments tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.

@@

The various bounds presented here overlap to some extent. For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its uses. For example, problems that lead to few successful matches but many type arguments tuples may benefit from a limit on the number of new type arguments tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.

@@

The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments tuples are generated respectively for each symbol at each iteration. These options are closely linked and must be set together if they are to be relevant.

During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments tuples is determined by this formula:
\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \cdot \text{number of type arguments})) \)
